---
title: "Project Step 2"
author: "Sebastien Montgrain"
format: 
  html:
    theme: minty
    code-tools: true
self-contained: true
editor: source
embed-resources: true
execute: 
  echo: true
  error: true
---

```{r}
#| message: False
library(tidyverse)
```

## Measles Cases Across The World

Reading in the data:

```{r}
#| message: False
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

### Context:

This data set is about measles and rubella cases around the world, provided by the World Health Organization on June 12th, 2025. It has data going back to 2012 of the monthly and yearly measles and rubella cases per country. It also has separate columns for suspected cases and confirmed cases, and some other classifications of cases.



### How do the trends in the proportions of lab-confirmed measles and rubella cases compare between 2012 and 2025?
Involving function and iteration from modified variable columns


```{r}
get_yearly_proportions <- function(y){
  cases_year |>
    filter(year == y) |>
    summarize(year = y,
          total = sum(measles_lab_confirmed + rubella_lab_confirmed),
          measles_prop = sum(measles_lab_confirmed/total),
          rubella_prop = sum(rubella_lab_confirmed/total)) |>
    select(year, measles_prop, rubella_prop)
}
```

```{r}
cases_year |>
  distinct(year) |>
  pull(year) |>
  map_dfr(get_yearly_proportions)
```
Results:
Measles have always been more common than Rubella, but the proportions seem to fluctuate over time.

### Which month typically has the most measles cases?
Modified Data, no iterations or custom functions


```{r}
Case_Ranking_By_Months <- cases_month |>
  group_by(month) |>
  summarise(total_measles = sum(
    measles_total,
    na.rm = TRUE)) |>
  mutate(rank = rank(
    -total_measles),
    month = month.name[month]) |>
  arrange(rank)

Case_Ranking_By_Months
```

Results: March has the most measles cases and September has the least.

### How do measles total and rubella total differ by region?
Involving function and iteration from existing variable columns

```{r}

summary_diff <- function(measles, rubella) {
  data.frame(
    m_mean = mean(as.numeric(measles)),
    r_mean = mean(as.numeric(rubella)),
    differ = mean(as.numeric(measles)) - mean(as.numeric(rubella))
  )
}
```

```{r}
region_diff <- cases_year |>
  group_by(region, year) |>
  summarise(summary_diff(measles_total, rubella_total), .groups = "drop")

region_diff

```
Likely need to visualize results to see patterns.

### Is there a relationship between rubella cases and measles cases?
Existing Data, there is a custom function & Iteration


```{r}
Corr_Measles_Rubella <- function(r) {
  cases_year |>
    filter(region == r) |>
    summarise(
      correlation = cor(
        as.numeric(measles_total),
        as.numeric(rubella_total),
        use = "complete.obs")) |>
    pull(correlation)}

Region_Correlation <- cases_year |>
  distinct(region) |>
  mutate(
    correlation = map_dbl(region, Corr_Measles_Rubella)
    ) |>
  arrange(desc(correlation))

Region_Correlation
```
Results: for some reason correlation is very high in South East Asia, and very weak in Africa, Europe, and the Americas.
